Hierarchical Embeddings for Disease-Signature Associations

### 2025-10-11 pyScripts/current copy.tex
Update

i% LaTeX template for Science journal paper
% Based on science_template.tex with adaptations for research content

@@

% Figure and Table labels in bold
\makeatletter
\renewcommand{\fnum@figure}{\textbf{Figure \thefigure}}
\renewcommand{\fnum@table}{\textbf{Table \thetable}}
\makeatother
\renewcommand\refname{References and Notes}
\makeatletter
\renewcommand{\fnum@figure}{\textbf{Figure \thefigure}}
\renewcommand{\fnum@table}{\textbf{Table \thetable}}
\makeatother
\usepackage{scicite}
\usepackage{url}
\usepackage{tikz}
%\newcommand{\aladyn}{\texttt{ALADYNOULLI}}
\newcommand{\aladyn}{\texttt{ALADYNOULLI}}
\usepackage{tikz}
\usepackage{amsmath}
\let\Bbbk\relax
\usepackage{amssymb}
\usepackage{geometry}
\geometry{margin=1in}
% Call the accompanying scicite.sty package.
% This formats citation numbers in Science style.
\usepackage{scicite}

@@

%%%%%%%%%%%%%%%% START OF MAIN TEXT %%%%%%%%%%%%%%%
\begin{document}

% Insert the title and author list
\maketitle

@@

\begin{abstract} \bfseries \boldmath
Understanding how disease patterns evolve over a lifetime remains a key challenge in medicine. While electronic health records provide rich longitudinal data, existing models typically analyze each disease in isolation, missing the complex interplay between multiple conditions and genetic factors. Here, we combine longitudinal health records with genetic data to model individual trajectories, using a novel latent dynamic Bayesian framework called $\aladyn$ that identifies latent disease signatures from longitudinal health records while modeling individual-specific trajectories. Applied across three biobanks with up to 50 years of follow-up, our model discovers clinically interpretable disease signatures that demonstrate remarkable consistency across diverse populations (79.2\% cross-cohort correspondence) and show strong genetic correlations, enabling both accurate prediction of patient risk and discovery of novel genetic associations. The model achieves dramatic improvements in disease prediction across 24 conditions (median dynamic AUC 0.85, gains up to 0.20), outperforming established clinical risk scores across diverse conditions. Furthermore, our signature-based approach identifies 150 novel genetic loci missed by single-disease GWAS, with multiple signatures showing strong genetic signals (cardiovascular h² = 0.041, musculoskeletal $h^{2}$ = 0.035). Critically, this unified modeling approach significantly improves predictive performance for multiple diseases while revealing distinct biological subtypes within traditional diagnostic categories—demonstrating substantial heterogeneity across diverse conditions including cancer, metabolic disorders, and psychiatric conditions, with Cohen's d effect sizes up to 3.87 for signature differences between patient clusters ($p \leq 1\times 10^{8}$ for 95\% of comparisons). In conclusion, $\aladyn$ combines genetics and longitudinal diagnosis to achieve both improved disease prediction and enhanced genetic discovery through a unified framework that captures the complex interplay between genetic predisposition and time-varying disease patterns.
\end{abstract}

\section*{Introduction}

>>>>

% LaTeX template for Science journal paper
% Based on science_template.tex with adaptations for research content

@@

% Figure and Table labels in bold
\makeatletter
\renewcommand{\fnum@figure}{\textbf{Figure \thefigure}}
\renewcommand{\fnum@table}{\textbf{Table \thetable}}
\makeatother
\renewcommand\refname{References and Notes}
\makeatletter
\renewcommand{\fnum@figure}{\textbf{Figure \thefigure}}
\renewcommand{\fnum@table}{\textbf{Table \thetable}}
\makeatother
\usepackage{scicite}
\usepackage{url}
\usepackage{tikz}
%\newcommand{\aladyn}{\texttt{ALADYNOULLI}}
\newcommand{\aladyn}{\texttt{ALADYNOULLI}}
\usepackage{tikz}
\usepackage{lineno}
\usepackage{amsmath}
\let\Bbbk\relax
\usepackage{amssymb}
\usepackage{geometry}
\geometry{margin=1in}
% Call the accompanying scicite.sty package.
% This formats citation numbers in Science style.
\usepackage{scicite}

@@

%%%%%%%%%%%%%%%% START OF MAIN TEXT %%%%%%%%%%%%%%%
\begin{document} 
\linenumbers
% Insert the title and author list
\maketitle

@@

\begin{abstract} \bfseries \boldmath
Understanding how disease patterns evolve over a lifetime remains a key challenge in medicine. While electronic health records provide rich longitudinal data, existing models typically analyze each disease in isolation, missing the complex interplay between multiple conditions and genetic factors. Here, we combine longitudinal health records with genetic data to model individual trajectories, using a novel dynamic Bayesian framework called $\aladyn$ that identifies latent disease signatures from longitudinal health records while modeling individual-specific trajectories. Applied across three biobanks with up to 50 years of follow-up, our model discovers clinically interpretable disease signatures that demonstrate remarkable consistency across diverse populations (79.2\% cross-cohort correspondence) and show strong genetic correlations, enabling both accurate prediction of patient risk and discovery of novel genetic associations. The model achieves dramatic improvements in disease prediction across 24 conditions (median dynamic AUC 0.85, gains up to 0.20), outperforming established clinical risk scores like PCE, PREVENT and GAIL. Furthermore, our signature-based approach identifies over 150 genetic loci - many missed by single-disease GWAS - with multiple signatures showing strong genetic signals (cardiovascular h² = 0.041, musculoskeletal $h^{2}$ = 0.035). Critically, this unified modeling approach significantly improves predictive performance for multiple diseases while revealing distinct biological subtypes within traditional diagnostic categories—demonstrating substantial heterogeneity across diverse conditions including cancer, metabolic disorders, and psychiatric conditions, with Cohen's d effect sizes up to 3.87 for signature differences between patient clusters ($p \leq 1\times 10^{8}$ for 95\% of comparisons). In conclusion, $\aladyn$ combines genetics and longitudinal diagnosis to achieve both improved disease prediction and enhanced genetic discovery through a unified framework that captures the complex interplay between genetic predisposition and time-varying disease patterns.
\end{abstract}

\newpage
\section*{Introduction}

### 2025-10-11 pyScripts/embddingslides.qmd
Update

:::: {.columns}

::: {.column width="50%"}
**Current approach:**
- Learn $\psi \in \mathbb{R}^{K \times D}$ directly
- $\psi_{k,d}$ = association strength between signature $k$ and disease $d$
:::

::: {.column width="50%"}
**Problems:**
- No semantic structure between diseases
- Difficult to generalize to new diseases
- No interpretable disease relationships
:::

::::

@@

:::: {.columns}

::: {.column width="50%"}
**Step 1: Disease and Signature Embeddings**
$$E_d \in \mathbb{R}^L \quad \text{(disease embeddings)}$$
$$E_k \in \mathbb{R}^L \quad \text{(signature embeddings)}$$
:::

@@

where $W_a \in \mathbb{R}^{L \times L}$ is learnable.
:::

::::

::: {.callout-note}
## Key Insight
Skip attention softmax and contextualization - use raw attention scores directly as ψ values.
:::

@@

| Variable | Shape | Description |
|----------|-------|-------------|
| $E_d$ | [D, L] | Disease embeddings |
| $E_k$ | [K, L] | Signature embeddings |
| $W_a$ | [L, L] | Attention interaction matrix |
| $\psi$ | [K, D] | Final association strengths |

@@

```{r embeddings_raw3}
ggplot(reshape2::melt(sim_data$E_k_true),aes(x=Var1,y=Var2,fill = value))+geom_tile()+labs(x="Signature",y="Embedding")+scale_fill_viridis()
```
## Disease Embeddings Visualization

@@

:::: {.columns}

::: {.column width="50%"}
**What $E_d$ tells us:**
Each disease is represented by an $L$-dimensional vector capturing its biological characteristics.

**What $E_k$ tells us:**
Each signature is represented by an $L$-dimensional vector capturing pathway characteristics.
:::

::: {.column width="50%"}
**What $W_a$ tells us:**
Learns which disease features should interact with which signature features.

**What $\psi_{k,d}$ tells us:**
Direct measure of how strongly disease $d$ associates with signature $k$ after $W_a$ transformation.
:::

::::

@@

::: {.callout-note}
## Key Components
```python
# Embeddings
self.disease_embeddings = nn.Embedding(D, L)
self.signature_embeddings = nn.Embedding(K, L)

@@

```python
def compute_psi_from_embeddings(self):
    E_d = self.disease_embeddings(torch.arange(D))  # [D, L]
    E_k = self.signature_embeddings(torch.arange(K))  # [K, L]

@@

:::: {.columns}

::: {.column width="33%"}
**Parameter Efficiency:**
- Direct: $O(D \times K)$
- Hierarchical: $O(D \times L + K \times L + L^2)$
- When $L \ll D$, much more efficient
:::

@@

::: {.column width="33%"}
**Regularization:**
- Embedding space provides smooth representations
- Attention encourages sparse associations
- Implicit regularization through shared representations
:::

::::

@@

:::: {.columns}

::: {.column width="50%"}
**Semantic Structure:**
- Diseases with similar embeddings have similar psi values
- Learned relationships are interpretable
- Attention weights reveal disease-signature associations
:::

::: {.column width="50%"}
**Model-Based Approach:**
- Maintains principled probabilistic framework
- Embeddings provide regularization
- Can incorporate external knowledge
:::

::::

---

@@

:::: {.columns}

::: {.column width="50%"}
**What We've Accomplished:**
- Replaced direct $\psi$ learning with hierarchical embeddings
- Used attention mechanism to relate diseases to signatures
- Created contextualized representations for each disease-signature pair
- Maintained model-based approach while adding interpretability
:::

::: {.column width="50%"}
**Next Steps:**
- Implement in our existing model (clust_gp)
- Experiment with different embedding dimensions
- Add medication effects
- Validate on our dataset
:::

::::

::: {.callout-tip}
## For Your Real Application
With D=350, K=20, use L=5-8 for optimal parameter efficiency while maintaining interpretability.
:::

>>>>

::::: columns
::: {.column width="50%"}
**Current approach:** - Learn $\psi \in \mathbb{R}^{K \times D}$ directly - $\psi_{k,d}$ = association strength between signature $k$ and disease $d$
:::

::: {.column width="50%"}
**Problems:** - No semantic structure between diseases - Difficult to generalize to new diseases - No interpretable disease relationships
:::
:::::

@@

::::: columns
::: {.column width="50%"}
**Step 1: Disease and Signature Embeddings** $$E_d \in \mathbb{R}^L \quad \text{(disease embeddings)}$$ $$E_k \in \mathbb{R}^L \quad \text{(signature embeddings)}$$
:::

@@

where $W_a \in \mathbb{R}^{L \times L}$ is learnable.
:::
:::::

::: callout-note
## Key Insight

Skip attention softmax and contextualization - use raw attention scores directly as ψ values.
:::

@@

| Variable | Shape    | Description                  |
|----------|----------|------------------------------|
| $E_d$    | \[D, L\] | Disease embeddings           |
| $E_k$    | \[K, L\] | Signature embeddings         |
| $W_a$    | \[L, L\] | Attention interaction matrix |
| $\psi$   | \[K, D\] | Final association strengths  |

@@

```{r embeddings_raw3}
ggplot(reshape2::melt(sim_data$E_k_true),aes(x=Var1,y=Var2,fill = value))+geom_tile()+labs(x="Signature",y="Embedding")+scale_fill_viridis()
```

## Disease Embeddings Visualization

@@

::::: columns
::: {.column width="50%"}
**What** $E_d$ tells us: Each disease is represented by an $L$-dimensional vector capturing its biological characteristics.

**What** $E_k$ tells us: Each signature is represented by an $L$-dimensional vector capturing pathway characteristics.
:::

::: {.column width="50%"}
**What** $W_a$ tells us: Learns which disease features should interact with which signature features.

**What** $\psi_{k,d}$ tells us: Direct measure of how strongly disease $d$ associates with signature $k$ after $W_a$ transformation.
:::
:::::

@@

::: callout-note
## Key Components

``` python
# Embeddings
self.disease_embeddings = nn.Embedding(D, L)
self.signature_embeddings = nn.Embedding(K, L)

@@

``` python
def compute_psi_from_embeddings(self):
    E_d = self.disease_embeddings(torch.arange(D))  # [D, L]
    E_k = self.signature_embeddings(torch.arange(K))  # [K, L]

@@

:::::: columns
::: {.column width="33%"}
**Parameter Efficiency:** - Direct: $O(D \times K)$ - Hierarchical: $O(D \times L + K \times L + L^2)$ - When $L \ll D$, much more efficient
:::

@@

::: {.column width="33%"}
**Regularization:** - Embedding space provides smooth representations - Attention encourages sparse associations - Implicit regularization through shared representations
:::
::::::

@@

::::: columns
::: {.column width="50%"}
**Semantic Structure:** - Diseases with similar embeddings have similar psi values - Learned relationships are interpretable - Attention weights reveal disease-signature associations
:::

::: {.column width="50%"}
**Model-Based Approach:** - Maintains principled probabilistic framework - Embeddings provide regularization - Can incorporate external knowledge
:::
:::::

------------------------------------------------------------------------

@@

::::: columns
::: {.column width="50%"}
**What We've Accomplished:** - Replaced direct $\psi$ learning with hierarchical embeddings - Used attention mechanism to relate diseases to signatures - Created contextualized representations for each disease-signature pair - Maintained model-based approach while adding interpretability
:::

::: {.column width="50%"}
**Next Steps:** - Implement in our existing model (clust_gp) - Experiment with different embedding dimensions - Add medication effects - Validate on our dataset
:::
:::::

::: callout-tip
## For Your Real Application

With D=350, K=20, use L=5-8 for optimal parameter efficiency while maintaining interpretability.
:::
